---
title: "Weight Gene Co-Expression Network Analysis (WGCNA)"
author: "Mariah Lee"
date: "Last Updated: 06.29.2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,      # Show code in the HTML (set to FALSE to hide code)
  warning = FALSE,  # Hide warnings in the output
  message = FALSE,  # Hide messages in the output
  fig.width = 7,    # Default figure width
  fig.height = 5,   # Default figure height
  fig.align = "center" # Center figures
)
```

## Overview

This R Markdown document initiates a co-expression analysis aimed at inferring the potential functions of long non-coding RNAs (lncRNAs) by identifying their expression relationships with protein-coding genes. By identifying highly co-expressed protein-coding genes, we can leverage established functional annotations (like Gene Ontology terms and biological pathways) of these coding partners to predict the roles of associated lncRNAs.

Be sure to change the following depending on your dataset:
- files to read in
- samples to be excluded (if any)
- soft power
- significant modules
- saved file names

## Load libraries

```{r libraries}
# List of required packages for co-expression and functional enrichment
cran_packages <- c("readr", "tidyverse", "knitr", "ggplot2", "ggrepel", "Hmisc",
                   "gridExtra", "devtools")
bioc_packages <- c("tximport", "edgeR", "DESeq2", "biomaRt", "clusterProfiler", 
                   "org.Hs.eg.db", "WGCNA")

# Function to check, install, and load packages
install_and_load <- function(pkg, bioc = FALSE) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    if (bioc) {
      if (!requireNamespace("BiocManager", quietly = TRUE))
        install.packages("BiocManager")
      BiocManager::install(pkg, ask = FALSE, update = FALSE)
    } else {
      install.packages(pkg)
    }
  }
  library(pkg, character.only = TRUE)
}

# Install and load CRAN packages
for (pkg in cran_packages) {
  install_and_load(pkg, bioc = FALSE)
}

# Install and load Bioconductor packages
for (pkg in bioc_packages) {
  install_and_load(pkg, bioc = TRUE)
}

# Install CorLevelPlot from GitHub
devtools::install_github("kevinblighe/CorLevelPlot")

library(CorLevelPlot)
allowWGCNAThreads()          # allow multi-threading
```

```{r data}
# Identify analysis
analysis_label <- "cd4_naive" # e.g. "cd4_all", "cd4_naive", or "cd4_memory"
# Load txi and metadata
txi_data <- readRDS(paste("../differential_expression/txi_", analysis_label, ".rds"))
metadata <- readRDS(paste("../differential_expression/meta", analysis_label, ".rds"))
```

## --- Outlier Detection for All CD4+ T Cells ---

```{r outliers}
# Get raw counts matrix
raw_counts <- txi_data$counts

# Check that all sample IDs are in the metadata
all(colnames(raw_counts) %in% metadata$SRR_ID)

# Detect outlier genes and samples
gsg <- goodSamplesGenes(t(raw_counts), verbose = 3)

# Report initial findings from goodSamplesGenes
print(paste("Original number of genes:", nrow(raw_counts)))
print(paste("Original number of samples:", ncol(raw_counts)))
print(paste("Genes flagged as problematic by goodSamplesGenes:", sum(!gsg$goodGenes)))
print(paste("Samples flagged as problematic by goodSamplesGenes:", sum(!gsg$goodSamples)))

# Remove genes/samples detected as outliers by goodSamplesGenes
if (!gsg$allOK) {
  gsg_counts <- raw_counts[gsg$goodGenes, gsg$goodSamples]
  # IMPORTANT: Update metadata to reflect samples removed by gsg
  metadata <- metadata %>% filter(SRR_ID %in% colnames(gsg_counts))
  # Re-order metadata to match the new gsg_counts_naive columns
  metadata <- metadata[match(colnames(gsg_counts), metadata$SRR_ID), ]

  print("Removed genes/samples flagged by goodSamplesGenes and updated metadata.")
} else {
  gsg_counts <- raw_counts # No changes if allOK
  # Ensure metadata is still in sync, even if no removals
  metadata <- metadata[match(colnames(gsg_counts), metadata$SRR_ID), ]
}
```

### Removing Outliers: Hierarchial Clustering

```{r clustering}
# Detect outlier samples using hierarchical clustering
htree <- hclust(dist(t(gsg_counts)), method = "average")

# Plot the hierarchical tree to visually inspect for additional outliers
sizeGrWindow(12, 9) # Adjust window size for plotting
par(cex = 0.8) # Adjust text size
par(mar = c(0, 5, 2, 0)) # Adjust margins
plot(htree, main = "Sample Clustering (Raw Counts)",
     sub = "", xlab = "", labels = colnames(gsg_counts), hang = -1)
```

### Removing Outliers: PCA

```{r pca}
# Detect outlier samples using PCA
pca <- prcomp(t(gsg_counts))
pca_dat <- as.data.frame(pca$x)
pca_var <- pca$sdev^2
pca_var_percent <- round(pca_var / sum(pca_var) * 100, digits = 2)

# Plot PCA to look for outliers in 2D space
ggplot(pca_dat, aes(PC1, PC2)) +
  geom_point() +
  geom_text(aes(label = rownames(pca_dat)), nudge_y = 0.5, size = 3) +
  labs(x = paste0('PC1: ', pca_var_percent[1], ' %'),
       y = paste0('PC2: ', pca_var_percent[2], ' %'),
       title = "PCA Plot for Outlier Detection") +
  theme_minimal()
```

### Removing Outlier Samples

```{r remove_samples}
# Define the list of outlier sample IDs you identified
samples_to_be_excluded <- c(
  "SRR22521436", "SRR22521557", "SRR22521556"
  )

# Remove selected outlier samples from the counts matrix
cleaned_counts <- gsg_counts[, !(colnames(gsg_counts) %in%
                                                   samples_to_be_excluded)]

# Remove selected outlier samples from the metadata dataframe
metadata_cleaned <- metadata %>%
  filter(!SRR_ID %in% samples_to_be_excluded)

# Verify the dimensions after removing these outliers
print(paste("Final number of genes:",
            nrow(cleaned_counts)))
print(paste("Final number of samples:",
            ncol(cleaned_counts)))
```

## --- Normalization for All CD4+ T Cells ---

```{r normalize}
# Create DESeq2 object
dds <- DESeqDataSetFromMatrix(
  countData = round(cleaned_counts),
  colData = metadata_cleaned,
  design = ~ 1 # Unsupervised design for WGCNA data preparation
)

# Filter out genes with very low counts
#    WGCNA recommends removing genes with counts < 15 in a high percentage of samples.
num_samples_cleaned <- ncol(cleaned_counts)
min_samples_with_count_threshold <- ceiling(num_samples_cleaned * 0.75) # At least 75% of samples

print(paste("Filtering genes with counts < 15 in more than", round(num_samples_cleaned * 0.25), 
            "samples (", round(num_samples_cleaned * 0.75), "% threshold)."))

dds_filtered <- dds[rowSums(counts(dds) >= 15) >= min_samples_with_count_threshold, ]

print(paste("Number of genes after filtering for low counts:", nrow(dds_filtered)))
```

### Variance Stabilization

```{r variance}
# Perform Variance Stabilizing Transformation (VST)
dds_final <- vst(dds_filtered)

# Extract normalized counts
norm_counts <- assay(dds_final) %>%
  t()

print(paste("Dimensions of normalized counts matrix (samples x genes):",
              nrow(norm_counts), "x", ncol(norm_counts)))
```

## --- Network Construction for All CD4+ T Cells ---

```{r soft_threshold}
# Soft-threshold test
power <- c(c(1:10), seq(from = 12, to = 50, by = 2))

# Pick soft threshold
sft <- pickSoftThreshold(norm_counts,
                         powerVector = power,
                         networkType = "signed", # To distinguish positive and negative correlations
                         verbose = 5) # verbose = 5 provides detailed output during computation

# Store the results of the soft threshold analysis
sft_data <- sft$fitIndices

# Visualization to pick the optimal soft power
#    Plot 1: Scale-free topology model fit (R^2) vs. Soft Threshold Power.
#    We look for the lowest power for which the R^2 value reaches a plateau, typically above 0.8.
topo <- ggplot(sft_data, aes(Power, SFT.R.sq, label = Power)) +
  geom_point() +
  geom_text(nudge_y = 0.1) +
  geom_hline(yintercept = 0.8, color = 'red', linetype = 'dashed') + # Target R^2 for scale-free topology
  labs(x = 'Soft Threshold Power', y = 'Scale-Free Topology Model Fit, signed R^2') +
  theme_classic() +
  ggtitle("Scale-Free Topology Fit")

# Plot 2: Mean Connectivity (k) vs. Soft Threshold Power.
# This plot shows how average connectivity decreases as power increases.
connect <- ggplot(sft_data, aes(Power, mean.k., label = Power)) +
  geom_point() +
  geom_text(nudge_y = 0.1) +
  labs(x = 'Soft Threshold Power', y = 'Mean Connectivity') +
  theme_classic() +
  ggtitle("Mean Connectivity")

# Arrange and display the plots
grid.arrange(topo, connect, nrow = 2)
```

### Build a Network

```{r network}
soft_power <- 16

# Convert matrix to numeric
norm_counts[] <- sapply(norm_counts, as.numeric)

# Temporarily store the original 'cor' function to avoid conflicts with WGCNA's 'cor'
temp_cor <- cor
cor <- WGCNA::cor

# Build the co-expression network and detect modules using blockwiseModules
bwnet <- blockwiseModules(norm_counts,
                          maxBlockSize = ncol(norm_counts), # Set to actual number of genes
                          TOMType = "signed", # Match networkType from pickSoftThreshold
                          power = soft_power, # Use the chosen soft threshold power
                          mergeCutHeight = 0.25,
                          numericLabels = FALSE, # Assign color names to modules
                          randomSeed = 1234, # For reproducibility
                          verbose = 3) # Provides progress updates

# Restore the original 'cor' function
cor <- temp_cor
```

## --- Module Eigengenes ---

This section extracts the module eigengenes (MEs) from the `blockwiseModules` output. Each ME represents the weighted average expression profile of all genes within a module and can be thought of as the "summary" of that module's activity. We also visualize the gene dendrogram alongside the detected module assignments, both before and after the module merging step.

```{r eigengenes}
# Extract module eigengenes from the blockwiseModules output
module_eigengenes <- bwnet$MEs

# Print out a preview of the module eigengenes (each column is an eigengene for a module)
print("Preview of Module Eigengenes:")
head(module_eigengenes)

# Get the number of genes assigned to each module color
print("Number of genes per module:")
table(bwnet$colors)

# Plot the gene dendrogram and the module colors before and after merging
sizeGrWindow(12, 9)
plotDendroAndColors(bwnet$dendrograms[[1]], # The first dendrogram
                    cbind(bwnet$unmergedColors, bwnet$colors),
                    c("Unmerged Modules", "Merged Modules"), # Labels for the color bars
                    dendroLabels = FALSE, # Do not plot gene names on the dendrogram
                    addGuide = TRUE, # Add horizontal lines to dendrogram
                    hang = 0.03,
                    guideHang = 0.05,
                    main = "Gene Dendrogram and Module Colors")

# The 'grey' module (if present) contains genes that do not strongly co-express with any other module.
```

## --- Relate Modules to Traits for All CD4+ T Cells ---

This crucial step correlates the module eigengenes (representing the overall expression of each module) with your relevant sample traits. This allows you to identify which modules are significantly associated with specific biological conditions or phenotypes. We'll use `Activation_Status` as the primary trait for this dataset.

```{r traits}
# Ensure 'Activation_Status' is a factor in your cleaned metadata.
metadata_cleaned$Activation_Status <- factor(metadata_cleaned$Activation_Status,
                                                     levels = c("resting", "activated"))

# Binarize the 'Activation_Status' categorical variable into numerical columns.
activation_status_bin <- binarizeCategoricalColumns(metadata_cleaned$Activation_Status,
                                                     includePairwise = FALSE,
                                                     includeLevelVsAll = TRUE,
                                                     minCount = 1 # Include all levels
                                                     )

# Create the final 'traits' dataframe for correlation.
traits <- data.frame(row.names = rownames(module_eigengenes))
traits <- cbind(traits, activation_status_bin)

head(traits)
```

```{r correlation}
# Define numbers of samples for correlation calculations
nSamples <- nrow(norm_counts)
nGenes <- ncol(norm_counts)

print(paste("Number of samples for correlation:", nSamples))
print(paste("Number of genes for correlation:", nGenes))


# Calculate the correlation between module eigengenes and traits
# 'use = 'p'' specifies Pearson correlation.
module_trait_corr <- cor(module_eigengenes, traits, use = 'p')
# Calculate the corresponding p-values for the correlations
module_trait_corr_pvals <- corPvalueStudent(module_trait_corr, nSamples)

print("Module-Trait Correlation Matrix:")
print(round(module_trait_corr, 2))
print("Module-Trait Correlation P-values:")
print(round(module_trait_corr_pvals, 3))
```

### Visualize module-trait association as a heatmap

```{r heatmap}
# Merge eigengenes with traits
heatmap_data <- merge(module_eigengenes, traits, by = 'row.names')
rownames(heatmap_data) <- heatmap_data$Row.names # Set sample IDs as row names
heatmap_data$Row.names <- NULL

# Identify the columns representing traits (for the X-axis) and module eigengenes (for the Y-axis)
traits_cols <- colnames(activation_status_bin)
me_cols <- colnames(module_eigengenes)

# Plot the heatmap using CorLevelPlot
CorLevelPlot(heatmap_data,
             x = traits_cols,
             y = me_cols,
             col = c("blue1", "skyblue", "white", "pink", "red"), # Custom color scale
             main = "Module-Trait Relationships")
```

### Create a mapping dataframe to easily find which gene belongs to which module

```{r map}
module_gene_mapping <- as.data.frame(bwnet$colors)
colnames(module_gene_mapping) <- "ModuleColor"
module_gene_mapping$Gene <- rownames(module_gene_mapping)

head(module_gene_mapping)
```

## --- Intramodular analysis: Identifying driver genes ---

```{r driver_genes}
# Calculate the Module Membership (kME) and associated p-values
# This measures how correlated each gene's expression is with the eigengene of each module.
module_membership <- cor(norm_counts, module_eigengenes, use = 'p')
module_membership_pvals <- corPvalueStudent(module_membership, nSamples)

print("Preview of Module Membership (kME) Matrix (Genes x Modules):")
print(head(module_membership[, 1:min(5, ncol(module_membership))]))
print("Preview of Module Membership P-values (Genes x Modules):")
print(head(module_membership_pvals[, 1:min(5, ncol(module_membership_pvals))]))
```

## Calculate Gene Significance (GS) and associated p-values for a specific trait

```{r gene_significance}
gene_signif <- cor(norm_counts, traits$data.activated.vs.all, use = 'p')
gene_signif_pvals <- corPvalueStudent(gene_signif, nSamples)

# Convert to data frame and arrange by p-value to see most significant genes
gene_signif_df <- as.data.frame(gene_signif)
colnames(gene_signif_df) <- "GS_Activation_Status.Activated"
gene_signif_df$Gene <- rownames(gene_signif_df)
gene_signif_df$PValue <- gene_signif_pvals

print(gene_signif_df %>%
        arrange(PValue) %>%
        head(25))
```

## --- Extract and save Hub Genes for significant modules ---

```{r complete_dataframe}
all_gene_info <- module_gene_mapping

# Add kME values and rename columns
kME_df <- as.data.frame(module_membership)
colnames(kME_df) <- paste0("kME_", gsub("ME", "", colnames(kME_df)))
kME_df$Gene <- rownames(kME_df)

# Add kME p-values
p_kME_df <- as.data.frame(module_membership_pvals)
colnames(p_kME_df) <- paste0("p.kME_", gsub("ME", "", colnames(p_kME_df)))
p_kME_df$Gene <- rownames(p_kME_df)

# Add Gene Significance (GS) values and their p-values
gene_level_data <- all_gene_info %>%
  left_join(kME_df, by = "Gene") %>%
  left_join(p_kME_df, by = "Gene") %>%
  left_join(gene_signif_df, by = "Gene")

save_filename <- paste0("gene_level_data_", analysis_label, ".rds")
saveRDS(gene_level_data, file = save_filename)

print(head(gene_level_data))
```

```{r hub_genes}
# Define your significant modules based on your module-trait heatmap and p-values
significant_modules <- c("turquoise", "brown")

# Define your thresholds for identifying hub genes (based only on kME)
kme_threshold <- 0.7
kme_pvalue_threshold <- 0.05

for (module_color in significant_modules) {
  print(paste("--- Processing module:", module_color, "---"))

  # Get all genes assigned to the current module
  current_module_genes_all_info <- gene_level_data %>%
    filter(ModuleColor == module_color)

  if (nrow(current_module_genes_all_info) == 0) {
    print(paste("No genes found for module:", module_color, ". Skipping."))
    next
  }

  print(paste("Total genes in", module_color, "module:", nrow(current_module_genes_all_info)))

  # Dynamically select the correct kME and p.kME columns for filtering
  kme_col <- paste0("kME_", module_color)
  p_kme_col <- paste0("p.kME_", module_color)

  # Check if the specific kME column exists for this module
  if (!kme_col %in% colnames(current_module_genes_all_info)) {
    print(paste("Warning: kME column", kme_col, "not found. Cannot filter by kME for this module."))
    next
  }

  # Filter for Hub Genes: high kME and low p-value
  hub_genes_in_module_df <- current_module_genes_all_info %>%
    filter(
      abs(.data[[kme_col]]) > kme_threshold,
      .data[[p_kme_col]] < kme_pvalue_threshold
    ) %>%
    # Select specific columns to save, including the relevant kME and p-value
    dplyr::select(Gene, ModuleColor, all_of(kme_col), all_of(p_kme_col),
                  GS_Activation_Status.Activated, PValue)

  print(paste("Number of Hub Genes in", module_color, "module:", nrow(hub_genes_in_module_df)))

  # Save the hub gene list to a CSV file
  if (nrow(hub_genes_in_module_df) > 0) {
    write.csv(hub_genes_in_module_df,
              file = paste0("HubGenes_", module_color, "_", analysis_label, ".csv"),
              row.names = FALSE)
    print(paste("Saved Hub Genes to:", 
                paste0("HubGenes_", module_color, "_", analysis_label, ".csv")))
  } else {
    print(paste("No Hub Genes found for", module_color, "with current thresholds."))
  }

  # Optional: Also save a simple list of ALL genes in the module (for broader functional enrichment)
  # This list would contain all genes assigned to the module, regardless of kME or GS.
  # If you want to do enrichment on the entire module first, then this list is what you'd use.
  write.table(current_module_genes_all_info$Gene,
              file = paste0("AllGenesInModule_", module_color, "_", analysis_label, ".tsv"),
              row.names = FALSE, col.names = FALSE, quote = FALSE)
  print(paste("Saved all genes in module to:", paste0("AllGenesInModule_", module_color,
                                                      "_", analysis_label, ".tsv")))
}
```

## Version Control

```{r versions}
# Get loaded packages and versions
packages <- sessionInfo()$otherPkgs
pkg_info <- tibble::tibble(
  Package = names(packages),
  Version = sapply(packages, function(pkg) pkg$Version)
)

# Save to text file
readr::write_tsv(pkg_info, "package_versions.txt")
```
