---
title: "Differential Expression in CD4+ T-cells - Activated vs Resting"
author: "Mariah Lee"
date: "Last Updated: 02.06.2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,      # Show code in the HTML (set to FALSE to hide code)
  warning = FALSE,  # Hide warnings in the output
  message = FALSE,  # Hide messages in the output
  fig.width = 7,    # Default figure width
  fig.height = 5,   # Default figure height
  fig.align = "center" # Center figures
)
```

## Overview

This document outlines a reproducible workflow for differential expression (DE) analysis comparing activated and resting CD4+ T cells. While the statistical modeling is performed on the whole transcriptome to ensure robust normalization and dispersion estimation, the downstream interpretation and visualization focus specifically on long non-coding RNAs (lncRNAs).

This analysis uses the following Bioconductor packages:

 - `tximport`: To import transcript-level quanitification from Salmon and summarize to the gene level, correcting for potential changes in isoform usage.

 - `DESeq2`: Used for the core differential expression analysis, utilizing a negative binomial generalized linear model to handle biological variability and low counts.
 
 - `biomaRt`: To map Ensembl identifiers to gene symbols and annotate gene biotypes (e.g., distinguishing lncRNAs from protein-coding genes).
 
 - `tidyverse`: For data manipulation and genersting publication-ready visualizations (Volcano plots, Boxplots).

Key steps in this workflow:

1. Import transcript-level quantification data from Salmon.
2. Annotate genes using Ensembl data.
3. Quality Control (QC): Assess sample similarity using PCA and check distribution using VST-normalized boxplots.
4. Filtering: Remove low-count genes to increase statistical power.
5. Differential Expression: Run the standard DESeq2 pipeline.
6. Refinement: Apply LFC Shrinkage to remove noise from low-count genes and validate using MA Plots.
7. Sanity Checks: Verify the analysis by plotting known T-cell activation markers (CD69, IL2RA, etc.).
8. Visualization: Generate Volcano plots to identify significant lncRNA candidates.
 
## Load libraries

```{r libraries}
# CRAN packages
library(tidyverse) # Includes ggplot2, readr, dplyr, etc.
library(knitr)
library(ggrepel)

# Bioconductor packages
library(tximport)
library(edgeR)
library(DESeq2)
library(biomaRt)
```

## Detect and List All Sample Quantification Files

This code chunk scans the `results/salmon` directory, identifies all sample subfolders (based on their naming pattern), and constructs a named vector of paths to each sample’s quant.sf file. This ensures the analysis will automatically include all current and future samples without manual updates.

```{r samples, echo=FALSE}
# List all entries in the results/salmon directory
all_entries <- list.dirs("../results/salmon", full.names = FALSE, recursive = FALSE)

# Filter to keep only sample folders (exclude known non-sample files/folders)
# You may want to exclude folders like 'deseq2_qc' or any that do not match the sample naming pattern
sample_folders <- all_entries[grepl("^SRR", all_entries)]

# Construct full paths to quant.sf for each sample
files <- file.path("../results/salmon", sample_folders, "quant.sf")
names(files) <- sample_folders

# Check result
head(files)
```

## Import Quantification Data and Summarize to Gene Level

This step imports transcript-level quantification files from Salmon for all detected samples, then uses the existing transcript-to-gene mapping file (tx2gene.tsv) to summarize the data at the gene level. The resulting `txi` objects contains estimated counts, abundances, and effective lengths, and serves as the direct input for downstream differential expression analysis.

```{r tximport}
# Read the transcript-to-gene mapping file
tx2gene <- readr::read_tsv("../results/salmon/tx2gene.tsv", col_types = "cc")

# Import quantification data and summarize to gene level
txi <- tximport(files, type = "salmon", tx2gene = tx2gene, countsFromAbundance = "no")

# Inspect the structure of the imported data
str(txi)
```

## Extract Gene-Level Counts for Visualization

While DESeq2 operates directly on the `txi` object, we extract the gene-level count matrix her for quality control purposes. This matrix allows for rapid inspection of library sizes and the generation of diagnostic plots, such as bloxplots and MDS, to assess sample quality before statistical testing.

```{r extract}
# Extract the counts matrix from tximport output
counts <- txi$counts
counts[1:5, 1:5]
```

## Load and Prepare Sample Metadata

This step reads the metadata CSV file into R, sets the sample IDs as row names for easy matching with quantification data, and checks that all samples in the expression matrix have corresponding metadata. This ensures the downstream analyses and visualizations can incorporate sample attributes like treatment, age, and sex.

```{r metadata}
# Load the metadata CSV file
metadata <- readr::read_csv("../metadata/sample_metadata.csv")

# Check that all sample names in the counts matrix are present in the metadata
all(colnames(counts) %in% metadata$SRR_ID)
# Should return TRUE; if not, check for mismatches or typos

# View the first few rows of the metadata to confirm successful import
head(metadata)
```

This chunk processes the metadata to add biologically relevant columns for downstream subsetting and analysis.

```{r edit_metadata}
# Add Cell_Type by extracting 'CD4' or 'CD8' from Biosample_term_name
metadata$Cell_Type <- str_extract(metadata$Biosample_term_name, "CD4|CD8")

# Add Cell_State based on keywords in Biosample_term_name
metadata$Cell_State <- "mature" # Default
metadata$Cell_State[grepl("naive", metadata$Biosample_term_name, ignore.case = TRUE)] <- "naive"
metadata$Cell_State[grepl("memory", metadata$Biosample_term_name, ignore.case = TRUE)] <- "memory"

# Rename 'Treatment' to 'Activation_Status' and recode values
metadata <- metadata %>%
  mutate(Activation_Status = recode(Treatment,
                                    "no_treatment" = "resting",
                                    "treatment" = "activated")) %>%
  dplyr::select(-Treatment)

# View the updated metadata
head(metadata)
```

Here we separate the data into appropriate categories based on cell state.

```{r subset}
# All CD4+ T cells (Activated vs Resting)
meta_cd4_all <- metadata %>%
  filter(Cell_Type == "CD4")

# Naive CD4+ T cells (Activated vs Resting)
meta_cd4_naive <- metadata %>%
  filter(Cell_Type == "CD4", Cell_State == "naive")

# Memory CD4+ T cells (Activated vs Resting)
meta_cd4_memory <- metadata %>%
  filter(Cell_Type == "CD4", Cell_State == "memory")


# Function to subset txi object based on sample IDs
subset_txi <- function(txi, sample_ids) {
  txi_subset <- list(
    counts = txi$counts[, sample_ids, drop=FALSE],
    abundance = txi$abundance[, sample_ids, drop=FALSE],
    length = txi$length[, sample_ids, drop=FALSE],
    countsFromAbundance = txi$countsFromAbundance
  )
}

# Subset txi for each comparison
txi_cd4_all    <- subset_txi(txi, meta_cd4_all$SRR_ID)
txi_cd4_naive  <- subset_txi(txi, meta_cd4_naive$SRR_ID)
txi_cd4_memory <- subset_txi(txi, meta_cd4_memory$SRR_ID)

# Save each subset
saveRDS(txi_cd4_all, "txi_cd4_all.rds")
saveRDS(meta_cd4_all, "meta_cd4_all.rds")

saveRDS(txi_cd4_naive, "txi_cd4_naive.rds")
saveRDS(meta_cd4_naive, "meta_cd4_naive.rds")

saveRDS(txi_cd4_memory, "txi_cd4_memory.rds")
saveRDS(meta_cd4_memory, "meta_cd4_memory.rds")
```

## Visualizing Gene Expression Data

Before starting differential expression analysis, it’s important to visualize the gene expression data to assess sample quality and relationships. We use Variance Stabilizing Transformation (VST) to normalize counts, which stabilizes variance across the mean, making it ideal for visualization (boxplots/PCA).

### Box Plot of VST-Transformed Counts

This plot shows the distribution of normalized counts. VST makes the data homoscedastic (constant variance), allowing us to fairly compare distributions across samples and identify technical outliers.

```{r boxplot}
# Create a temporary DESeqDataSet for QC purposes
dds_qc <- DESeqDataSetFromTximport(txi, colData = metadata, design = ~ 1)

# Perform Variance Stabilizing Transformation (VST)
vsd <- vst(dds_qc, blind = TRUE)

# Extract the matrix of transformed counts
vst_mat <- assay(vsd)

# Convert to long format for ggplot2
vst_long <- as.data.frame(vst_mat) %>%
  rownames_to_column("Gene") %>%
  pivot_longer(cols = -Gene, names_to = "Sample", values_to = "VST_Count")

# Box plot
p_box <- ggplot(vst_long, aes(x = Sample, y = VST_Count)) +
  geom_boxplot() +
  labs(title = "Distribution of VST-Transformed Counts",
       y = "VST Normalized Counts", x = "Sample") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Display and Save
print(p_box)
ggsave("VST_Distribution_Boxplot.png", plot = p_box, width = 8, height = 6)
```

### PCA Plot of Samples

Principal Component Analysis (PCA) visualizes the variance in the dataset. We use the VST-normalized data to ensure that highly expressed genes don't dominate the clustering purely due to magnitude.

```{r pca}
# Generate PCA Data
pcaData <- plotPCA(vsd, intgroup = "Activation_Status", returnData = TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))

# Plot using ggplot2
p_pca <- ggplot(pcaData, aes(PC1, PC2, color = Activation_Status, shape = Cell_Type)) +
  geom_point(size = 3) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) + 
  coord_fixed() +
  ggtitle("PCA of CD4+ T Cell Subsets") +
  theme_bw()

# Display and Save
print(p_pca)
ggsave("PCA_Plot.png", plot = p_pca, width = 8, height = 6)
```

## Add Gene Names/Symbols to Counts

```{r gene_symbols}
# Connect to Ensembl Mart
ensembl <- useEnsembl(biomart = "ensembl", 
                      dataset = "hsapiens_gene_ensembl", 
                      host = "https://www.ensembl.org")

# Get gene IDs from your data
gene_ids_with_versions <- rownames(txi$counts)

# Remove the ".#" version suffix from the Ensembl IDs
gene_ids <- sub("\\..*", "", gene_ids_with_versions)

gene_annotations <- getBM(
  attributes = c('ensembl_gene_id', 'hgnc_symbol', 'description', 'gene_biotype'),
  filters = 'ensembl_gene_id',
  values = gene_ids,
  mart = ensembl
)

# Identify Ensembl IDs that map to more than one unique HGNC symbol
multi_symbol_mappings <- gene_annotations %>%
  filter(hgnc_symbol != "") %>% # Only consider entries with an actual symbol
  group_by(ensembl_gene_id) %>%
  summarise(
    num_unique_symbols = n_distinct(hgnc_symbol),
    symbols = paste(sort(unique(hgnc_symbol)), collapse = "; ") # For inspection
  ) %>%
  filter(num_unique_symbols > 1)
```

## Create a map for gene symbols

```{r create_symbol_map}
# Add full gene IDs with versions to gene_annotations
gene_annotations$gene_id_with_version <- gene_ids_with_versions[
 match(gene_annotations$ensembl_gene_id, gene_ids)
]

# Create a mapping table: gene_id_with_version -> symbol/biotype
gene_symbol_map <- gene_annotations %>%
 dplyr::select(gene_id_with_version, hgnc_symbol, gene_biotype) %>%
 mutate(symbol_or_id = ifelse(hgnc_symbol == "", gene_id_with_version, hgnc_symbol)) %>%
 distinct(gene_id_with_version, .keep_all = TRUE)
```

## Define Long Non-Coding RNA (lncRNA) Subset

Although the statistical analysis is performed on the entire transcriptome to maintain rigorous normalization, we define a specific list of lncRNAs here. This subset will be used to filter and highlight relevant candidates in downstream visualizations and result tables.

```{r lncRNAs}
# Filter the gene_annotations to identify genes with 'lncRNA' biotype
lncrna_list <- gene_annotations %>%
  filter(gene_biotype == "lncRNA")

# Display the first few entries and the dimensions of the lncRNA list
cat("First few entries of the lncRNA list:\n")
print(head(lncrna_list))
cat("\nDimensions of the lncRNA list (rows, columns): ", dim(lncrna_list), "\n")
```

## Create DESeq2 Objects

```{r objects}
dds_cd4_all <- DESeqDataSetFromTximport(
  txi = txi_cd4_all,
  colData = meta_cd4_all,
  design = ~ Activation_Status
)

dds_cd4_naive <- DESeqDataSetFromTximport(
  txi = txi_cd4_naive,
  colData = meta_cd4_naive,
  design = ~ Activation_Status
)

dds_cd4_memory <- DESeqDataSetFromTximport(
  txi = txi_cd4_memory,
  colData = meta_cd4_memory,
  design = ~ Activation_Status
)
```

## Filter Low Counts

Keep gene if it has >= 5 counts in at least 25% of the samples

```{r filter_low_counts}
filter_dds <- function(dds) {
  min_count <- 5
  min_samples <- ceiling(ncol(dds) / 4)
  
  keep <- rowSums(counts(dds) >= min_count) >= min_samples
  return(dds[keep,])
}

dds_cd4_all_filtered <- filter_dds(dds_cd4_all)
dds_cd4_naive_filtered <- filter_dds(dds_cd4_naive)
dds_cd4_memory_filtered <- filter_dds(dds_cd4_memory)

print(paste("Genes remaining in All CD4+ set:", nrow(dds_cd4_all_filtered)))
```

## Differential Expression Analysis: Activated vs Resting CD4⁺ T Cells

This section performs the core differential expression analysis using the `DESeq2` package. We define a function to run the differential expression pipeline, apply Log Fold Change (LFC) shrinkage to reduce noise in low-count genes, and generate MA plots for quality control.

```{r de_analysis}
# Function to run DESeq2, Shrinkage, and MA Plot
run_deseq_analysis <- function(dds_obj, analysis_name) {
  
  # Ensure "resting" is the reference level
  dds_obj$Activation_Status <- factor(dds_obj$Activation_Status, 
                                      levels = c("resting", "activated"))
  
  # Perform differential expression
  dds <- DESeq(dds_obj)
  
  # Extract and Save Normalized Counts
  # Clean the name for the filename (replace spaces with underscores)
  clean_name <- gsub(" ", "_", analysis_name)
  norm_counts <- counts(dds, normalized = TRUE)
  write.csv(norm_counts, paste0("normalized_counts_", clean_name, ".csv"))
  
  # Extract Results
  res <- results(dds, contrast = c("Activation_Status", "activated", "resting"))
  
  # Apply LFC Shrinkage (type="normal" is built-in; use "apeglm" if installed for better accuracy)
  res_shrunk <- lfcShrink(dds, 
                          contrast = c("Activation_Status", "activated", "resting"), 
                          res = res, type = "normal")
  
  # Save MA Plot to File
  png(paste0("MA_Plot_", clean_name, ".png"), width = 800, height = 600)
  plotMA(res_shrunk, ylim = c(-5, 5), main = paste("MA Plot:", analysis_name))
  dev.off()
  
  # Display MA Plot in Document
  plotMA(res_shrunk, ylim = c(-5, 5), main = paste("MA Plot:", analysis_name))
  
  return(res_shrunk)
}

# --- Execute Analysis for All Subsets ---
print("--- Processing All CD4+ ---")
results_cd4_all <- run_deseq_analysis(dds_cd4_all_filtered, "All CD4+ T Cells")

print("--- Processing Naive ---")
results_cd4_naive <- run_deseq_analysis(dds_cd4_naive_filtered, "Naive CD4+ T Cells")

print("--- Processing Memory ---")
results_cd4_memory <- run_deseq_analysis(dds_cd4_memory_filtered, "Memory CD4+ T Cells")
```

## Identification of Differentially Expressed lncRNAs

After processing, here we save CSVs of the significant lncRNAs that are differentially expressed.

```{r identify_sig_lncrnas}
# Define function to extract significant lncRNAs
get_sig_lncrnas <- function(res_obj, symbol_map) {
  as.data.frame(res_obj) %>%
    rownames_to_column("gene_id") %>%
    left_join(symbol_map, by = c("gene_id" = "gene_id_with_version")) %>%
    filter(padj <= 0.05, abs(log2FoldChange) >= 1) %>%
    filter(gene_biotype == "lncRNA") %>%
    arrange(padj)
}

# Generate lists for each subset
sig_lncrnas_all    <- get_sig_lncrnas(results_cd4_all, gene_symbol_map)
sig_lncrnas_naive  <- get_sig_lncrnas(results_cd4_naive, gene_symbol_map)
sig_lncrnas_memory <- get_sig_lncrnas(results_cd4_memory, gene_symbol_map)

# Save the primary list to a CSV for external review
write_csv(sig_lncrnas_all, "significant_lncrnas_all_cd4.csv")
write_csv(sig_lncrnas_naive, "significant_lncrnas_naive_cd4.csv")
write_csv(sig_lncrnas_memory, "significant_lncrnas_memory_cd4.csv")

# Capture counts for the personal statement
all_count <- nrow(sig_lncrnas_all)

cat("Total significantly expressed lncRNAs in All CD4+ T-cells:", all_count, "\n")
```

## Visualization: Volcano Plots

These volcano plots visualize the differential gene expression results for Activated versus Resting CD4+ T cells. Positive LFC values mean the gene is more highly expressed in Activated CD4+ T cells compared to Resting CD4+ T cells. Negative LFC values mean the gene is more highly expressed in Resting CD4+ T cells compared to Activated CD4+ T cells.

```{r volcanos}
# Function to create Volcano Plot
plot_volcano <- function(res_obj, symbol_map, title_text) {
  
  # Prepare Data
  plot_data <- as.data.frame(res_obj) %>%
    rownames_to_column("gene_id") %>%
    left_join(symbol_map, by = c("gene_id" = "gene_id_with_version"))
  
  # Define Thresholds
  padj_threshold <- 0.05
  lfc_threshold <- 1
  
  # Categorize Significance
  plot_data <- plot_data %>%
    mutate(Significance = case_when(
      padj < padj_threshold & log2FoldChange > lfc_threshold ~ "Upregulated",
      padj < padj_threshold & log2FoldChange < -lfc_threshold ~ "Downregulated",
      TRUE ~ "Not Significant"
    ))
  
  plot_data$Significance <- factor(plot_data$Significance, 
                                   levels = c("Upregulated", "Downregulated", "Not Significant"))
  
  # Filter Labels (Significant lncRNAs only)
  labels_data <- plot_data %>%
    filter(Significance %in% c("Upregulated", "Downregulated")) %>%
    filter(gene_biotype == "lncRNA") %>%
    arrange(padj) %>%
    slice_head(n = 20)
  
  # Plot
  ggplot(plot_data, aes(x = log2FoldChange, y = -log10(padj))) +
    geom_point(aes(color = Significance), alpha = 0.8, size = 1) +
    scale_color_manual(values = c("Upregulated" = "red", 
                                  "Downregulated" = "blue", 
                                  "Not Significant" = "grey")) +
    geom_vline(xintercept = c(-lfc_threshold, lfc_threshold), 
               linetype = "dashed", color = "darkgreen") +
    geom_hline(yintercept = -log10(padj_threshold), 
               linetype = "dashed", color = "darkgreen") +
    geom_text_repel(data = labels_data, aes(label = symbol_or_id), 
                    size = 2.5, box.padding = 0.5, max.overlaps = Inf) +
    labs(title = title_text, x = "Log2 Fold Change (Shrunk)", 
         y = "-log10(Adjusted p-value)") +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"), 
          legend.title = element_blank())
}

# --- Generate Plots ---
# Note: fig.height ensures they display large enough in the HTML output
p1 <- plot_volcano(results_cd4_all, gene_symbol_map, "Activated vs Resting (All CD4+)")
p2 <- plot_volcano(results_cd4_naive, gene_symbol_map, "Activated vs Resting (Naive)")
p3 <- plot_volcano(results_cd4_memory, gene_symbol_map, "Activated vs Resting (Memory)")

# Save plots
ggsave("Volcano_All_CD4.png", p1, width = 7, height = 6)
ggsave("Volcano_Naive.png", p2, width = 7, height = 6)
ggsave("Volcano_Memory.png", p3, width = 7, height = 6)

# Display plots
p1
p2
p3
```

## Sanity Check: Activation Markers & Known LncRNAs

We verify the validity of the analysis by plotting the expression of well-established T-cell activation markers. We expect strong upregulation of early activation markers such as **CD69** and **IL2RA** in the activated condition compared to resting. LncRNAs that are known activation markers are also included.

```{r sanity_check}
# Function to plot specific markers
plot_marker_gene <- function(dds_obj, symbol, title_prefix) {
  
  # Map Symbol to Ensembl ID
  target_id <- gene_symbol_map %>% 
    filter(hgnc_symbol == symbol) %>% 
    pull(gene_id_with_version)
  
  # Plot if found
  if (length(target_id) > 0 && target_id %in% rownames(dds_obj)) {
    d <- plotCounts(dds_obj, gene = target_id, intgroup = "Activation_Status", returnData = TRUE)
    
    p <- ggplot(d, aes(x = Activation_Status, y = count, fill = Activation_Status)) +
      geom_boxplot(alpha = 0.8, outlier.shape = NA) +
      geom_point(position = position_jitter(w = 0.1, h = 0), size = 2) +
      scale_y_log10() +
      scale_fill_manual(values = c("activated" = "red", "resting" = "blue")) +
      labs(title = paste(title_prefix, ":", symbol), y = "Normalized Counts") +
      theme_bw() +
      theme(legend.position = "none", plot.title = element_text(hjust = 0.5))
    
    print(p)
    ggsave(paste0("SanityCheck_", title_prefix, "_", symbol, ".png"), p, width = 5, height = 4)
  } else {
    print(paste("Marker", symbol, "not found (low counts or ID mismatch)."))
  }
}

# --- General Markers & LncRNAs (Check in All CD4+) ---
# CD69/IL2RA = Gold standard activation
# NEAT1/MALAT1 = Known activation-associated lncRNAs
general_set <- c("CD69", "IL2RA", "CTLA4", "NEAT1", "MALAT1", "LINC00426", "LINC00152")

print("--- Checking General Markers & LncRNAs ---")
for (gene in general_set) {
  plot_marker_gene(dds_cd4_all_filtered, gene, "All_CD4")
}

# --- Naive-Biased Response (Check in Naive) ---
# Early immediate genes often spike highest here
print("--- Checking Naive Response ---")
for (gene in c("FOS", "JUN")) {
  plot_marker_gene(dds_cd4_naive_filtered, gene, "Naive")
}

# --- Effector Response (Check in Memory) ---
# Memory cells should rapidly produce effector cytokines
print("--- Checking Memory Effector Response ---")
for (gene in c("IFNG", "GZMB")) {
  plot_marker_gene(dds_cd4_memory_filtered, gene, "Memory")
}
```

## Saving the Results

```{r save}
# Function to filter and save significant results
save_results <- function(res_obj, filename, symbol_map) {
  df <- as.data.frame(res_obj) %>%
    rownames_to_column("gene_id") %>%
    left_join(symbol_map, by = c("gene_id" = "gene_id_with_version")) %>%
    filter(padj <= 0.05, abs(log2FoldChange) >= 1) %>%
    arrange(padj)
  
  write_tsv(df, filename)
}

save_results(results_cd4_all, "results_cd4_all_significant.tsv", gene_symbol_map)
save_results(results_cd4_naive, "results_cd4_naive_significant.tsv", gene_symbol_map)
save_results(results_cd4_memory, "results_cd4_memory_significant.tsv", gene_symbol_map)
```

## Version Control

```{r versions}
# Get loaded packages and versions
packages <- sessionInfo()$otherPkgs
pkg_info <- tibble::tibble(
  Package = names(packages),
  Version = sapply(packages, function(pkg) pkg$Version)
)

# Save to text file
readr::write_tsv(pkg_info, "package_versions.txt")
```
