---
title: "Functional Enrichment"
author: "Mariah Lee"
date: 'Last Updated: 02.11.2026'
output: html_document
params:
  # Select Analysis (Must match names in configs below)
  analysis: "cd4_all"
  
  # Select Specific Module to Enrich (e.g., "brown")
  target_module: "blue"

  # Reference Configs (Copied from WGCNA script for consistency)
  configs:
    value:
      cd4_all:
        sig_modules: ["turquoise", "darkred", "cyan", "yellow", "black", "brown", "lightyellow", "blue"]
      
      cd4_naive:
        sig_modules: ["pink", "brown", "turquoise", "blue", "green"]
      
      cd4_memory:
        sig_modules: [] # No significant modules analyzed
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,      # Show code in the HTML (set to FALSE to hide code)
  warning = FALSE,  # Hide warnings in the output
  message = FALSE,  # Hide messages in the output
  fig.width = 7,    # Default figure width
  fig.height = 5,   # Default figure height
  fig.align = "center" # Center figures
)
```

## Overview

This document performs functional and pathway enrichment analyses on gene lists (hub genes and whole modules) identified from WGCNA. It connects the "color" of a module to biological functions (GO terms and KEGG pathways).

## Load libraries

```{r libraries}
# Data Manipulation & Plotting
library(tidyverse)
library(knitr)

# Enrichment Analysis
library(clusterProfiler)
library(org.Hs.eg.db)
library(DOSE)
library(enrichplot)
library(pathview) # Optional: Only load if you plan to visualize KEGG maps
```

## Define Analysis Parameters

We extract the specific settings for this run from the YAML header.

```{r parameters}
# Get Parameters
analysis_label <- params$analysis
module_color <- params$target_module

# Validation
# Check if the chosen module is actually in the config list for this analysis
valid_modules <- params$configs[[analysis_label]]$sig_modules
if (!(module_color %in% valid_modules)) {
  warning(paste("Module", module_color, "is not listed as significant in the config for", analysis_label))
}

# Create Output Directory
results_dir <- file.path("enrichment_results", analysis_label)
if (!dir.exists(results_dir)) dir.create(results_dir, recursive = TRUE)

print(paste("Running Enrichment for Analysis:", analysis_label))
print(paste("Target Module:", module_color))
```

## Load Gene Lists

We load the specific hub genes and module membership lists generated by the WGCNA script.

```{r all_genes}
# Define Paths
base_path <- file.path("..", "co-expression")

hub_genes_file <- file.path(base_path, 
                            paste0("HubGenes_", module_color, "_", analysis_label, ".csv"))
all_genes_file <- file.path(base_path, 
                            paste0("AllGenesInModule_", module_color, "_", analysis_label, ".tsv"))
background_file <- file.path(base_path, 
                             paste0("gene_level_data_", analysis_label, ".rds"))

# Load Hub Genes
if (file.exists(hub_genes_file)) {
  hub_genes_df <- read.csv(hub_genes_file)
  gene_list_hub <- hub_genes_df$Gene
  print(paste("Loaded", length(gene_list_hub), "Hub Genes."))
} else {
  stop(paste("Hub gene file not found:", hub_genes_file))
}

# Load All Genes in Module
if (file.exists(all_genes_file)) {
  gene_list_all <- read.table(all_genes_file, header = FALSE, stringsAsFactors = FALSE)$V1
  print(paste("Loaded", length(gene_list_all), "total genes in module."))
} else {
  stop(paste("All genes file not found:", all_genes_file))
}

# Load Background Universe (Crucial for Statistics)
if (file.exists(background_file)) {
  bg_data <- readRDS(background_file)
  background_genes <- as.character(bg_data$Gene)
  print(paste("Background universe size:", length(background_genes)))
} else {
  stop(paste("Background data not found:", background_file))
}
```

## Gene ID Conversion

This step converts the gene lists (hub genes, all genes in module, and background genes) from gene symbols to ENTREZ IDs. This is often necessary as many `clusterProfiler` functions require ENTREZ IDs for proper mapping to biological databases.

```{r convert}
# Helper function to clean IDs and convert to Entrez
convert_to_entrez <- function(gene_list) {
  # 1. Strip version numbers (e.g., ENSG00000123.4 -> ENSG00000123)
  clean_genes <- sub("\\..*", "", gene_list)
  
  # 2. Convert ENSEMBL to ENTREZID
  bitr(clean_genes, 
       fromType = "ENSEMBL", 
       toType = "ENTREZID", 
       OrgDb = org.Hs.eg.db, 
       drop = TRUE)
}

hub_entrez <- convert_to_entrez(gene_list_hub)
all_entrez <- convert_to_entrez(gene_list_all)
bg_entrez  <- convert_to_entrez(background_genes)

print(paste("Mapped Hub Genes:", nrow(hub_entrez)))
print(paste("Mapped Module Genes:", nrow(all_entrez)))
print(paste("Mapped Background:", nrow(bg_entrez)))
```

## Functional Enrichment Functions

Here reusable functions are defined to run enrichment and save results automatically.

```{r enrichment_functions}
# Function to run GO Enrichment (BP, MF, or CC)
run_go_enrichment <- function(genes, background, ontology, title_prefix) {
  
  print(paste("Running GO:", ontology, "for", title_prefix))
  
  ego <- enrichGO(gene          = genes,
                  universe      = background,
                  OrgDb         = org.Hs.eg.db,
                  keyType       = "ENTREZID",
                  ont           = ontology,
                  pAdjustMethod = "BH",
                  pvalueCutoff  = 0.05,
                  qvalueCutoff  = 0.05,
                  readable      = TRUE) # Converts IDs back to Symbols in result
  
  # Save and Plot if results exist
  if (!is.null(ego) && nrow(as.data.frame(ego)) > 0) {
    # File naming
    file_base <- paste0(analysis_label, "_", module_color, "_GO_", ontology, "_", title_prefix)
    
    # Save Table
    write.csv(as.data.frame(ego), file.path(results_dir, paste0(file_base, ".csv")))
    
    # Plot
    p <- dotplot(ego, showCategory = 15, title = paste("GO", ontology, ":", title_prefix))
    print(p)
    ggsave(file.path(results_dir, paste0(file_base, ".png")), p, width = 8, height = 6)
    
  } else {
    print(paste("No significant terms found for", ontology, "-", title_prefix))
  }
}

# Function to run KEGG Enrichment
run_kegg_enrichment <- function(genes, background, title_prefix) {
  
  print(paste("Running KEGG for", title_prefix))
  
  kegg <- enrichKEGG(gene         = genes,
                     universe     = background,
                     organism     = 'hsa',
                     pvalueCutoff = 0.05)
  
  if (!is.null(kegg) && nrow(as.data.frame(kegg)) > 0) {
    # File naming
    file_base <- paste0(analysis_label, "_", module_color, "_KEGG_", title_prefix)
    
    # Save Table
    write.csv(as.data.frame(kegg), file.path(results_dir, paste0(file_base, ".csv")))
    
    # Plot
    p <- dotplot(kegg, showCategory = 15, title = paste("KEGG:", title_prefix))
    print(p)
    ggsave(file.path(results_dir, paste0(file_base, ".png")), p, width = 8, height = 6)
    
  } else {
    print(paste("No significant KEGG pathways found for", title_prefix))
  }
}
```

## Run Enrichment Analysis

The functions are applied to both hub genes (high confidence drivers) and all genes (broad functional context).

```{r run_enrichment}
# 1. Hub Genes Analysis
print("--- Analyzing Hub Genes ---")
run_go_enrichment(hub_entrez$ENTREZID, bg_entrez$ENTREZID, "BP", "HubGenes")
run_go_enrichment(hub_entrez$ENTREZID, bg_entrez$ENTREZID, "MF", "HubGenes")
run_go_enrichment(hub_entrez$ENTREZID, bg_entrez$ENTREZID, "CC", "HubGenes")
run_kegg_enrichment(hub_entrez$ENTREZID, bg_entrez$ENTREZID, "HubGenes")

# 2. All Module Genes Analysis
print("--- Analyzing All Module Genes ---")
run_go_enrichment(all_entrez$ENTREZID, bg_entrez$ENTREZID, "BP", "AllGenes")
run_go_enrichment(all_entrez$ENTREZID, bg_entrez$ENTREZID, "MF", "AllGenes")
run_go_enrichment(all_entrez$ENTREZID, bg_entrez$ENTREZID, "CC", "AllGenes")
run_kegg_enrichment(all_entrez$ENTREZID, bg_entrez$ENTREZID, "AllGenes")
```

## Version Control

```{r versions}
# Save versions
readr::write_tsv(tibble(
  Package = names(sessionInfo()$otherPkgs),
  Version = sapply(sessionInfo()$otherPkgs, function(pkg) pkg$Version)
), "Enrichment_package_versions.txt")
```
