---
title: "Weighted Gene Co-Expression Network Analysis (WGCNA)"
author: "Mariah Lee"
date: "Last Updated: 02.10.2026"
output: html_document
params:
  # CHANGE THIS LINE ONLY: "cd4_all", "cd4_naive", or "cd4_memory"
  analysis: "cd4_all"
  
  # Settings Dictionary
  configs:
    value:
      cd4_all:
        soft_power: 14
        samples_to_remove: ["SRR22521436", "SRR22521557", "SRR22521556", "SRR22520975", "SRR22520976"]
        sig_modules: ["turquoise", "darkred", "cyan", "yellow", "black", "brown", "lightyellow", "blue"]
      
      cd4_naive:
        soft_power: 16
        samples_to_remove: ["SRR22521436", "SRR22521557", "SRR22521556"]
        sig_modules: ["pink", "brown", "turquoise", "blue", "green"]
      
      cd4_memory:
        soft_power: 18
        samples_to_remove: []
        sig_modules: ["blue", "turquoise"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,      # Show code in the HTML (set to FALSE to hide code)
  warning = FALSE,  # Hide warnings in the output
  message = FALSE,  # Hide messages in the output
  fig.width = 7,    # Default figure width
  fig.height = 5,   # Default figure height
  fig.align = "center" # Center figures
)
```

## Overview

This R Markdown document performs **Weighted Gene Co-Expression Network Analysis (WGCNA)** to identify clusters (modules) of highly correlated genes. The primary objective is to infer the biological functions of **long non-coding RNAs (lncRNAs)** by associating them with co-expressed protein-coding genes and specific T-cell activation states.

**Key Features:**
* **Parameterized Workflow:** Designed to be run dynamically for different cell subsets (All CD4+, Naive, Memory) by changing the `analysis` parameter in the YAML header.
* **Systems Biology:** Moves beyond differential expression to identify regulatory networks, correlated modules, and "hub" driver genes.

## Load libraries

```{r libraries}
# Standard Data & Plotting
library(tidyverse)
library(knitr)
library(ggrepel)
library(gridExtra)
library(Hmisc)

# Bioconductor / Genomics
library(tximport)
library(edgeR)
library(DESeq2)
library(biomaRt)
library(clusterProfiler)
library(org.Hs.eg.db)

# Network Analysis
library(WGCNA)

# Install CorLevelPlot from GitHub
#devtools::install_github("kevinblighe/CorLevelPlot")
library(CorLevelPlot) 

allowWGCNAThreads() # Enable multi-threading
```

```{r setup_config}
# Get the active analysis name from YAML
analysis_label <- params$analysis

# Dynamically extract the configuration for that specific analysis
# This uses the string "cd4_all" (or whatever you set) to grab the right list
current_config <- params$configs[[analysis_label]]

# Unpack variables for the rest of the script
soft_power <- current_config$soft_power
samples_to_be_excluded <- current_config$samples_to_remove
significant_modules <- current_config$sig_modules

# Handle empty list for samples_to_remove (converts NULL to empty vector)
if (is.null(samples_to_be_excluded)) samples_to_be_excluded <- character(0)

# Verification Output
print(paste("Running Analysis:", analysis_label))
print(paste("Soft Power:", soft_power))
print("Excluding Samples:"); print(samples_to_be_excluded)
```

```{r data}
# Load txi and metadata
txi_file <- file.path("..", "differential_expression", 
                      paste0("txi_", analysis_label, ".rds"))
meta_file <- file.path("..", "differential_expression", 
                       paste0("meta_", analysis_label, ".rds"))

txi_data <- readRDS(txi_file)
metadata <- readRDS(meta_file)
```

## Outlier Detection and Removal

```{r outliers}
# Get raw counts matrix
raw_counts <- txi_data$counts

# Check that all sample IDs are in the metadata
all(colnames(raw_counts) %in% metadata$SRR_ID)

# Detect outlier genes and samples
gsg <- goodSamplesGenes(t(raw_counts), verbose = 3)

# Report initial findings from goodSamplesGenes
print(paste("Original number of genes:", nrow(raw_counts)))
print(paste("Original number of samples:", ncol(raw_counts)))

# Remove genes/samples detected as outliers by goodSamplesGenes
if (!gsg$allOK) {
  # Identify specific removals
  removed_samples <- colnames(raw_counts)[!gsg$goodSamples]
  removed_genes <- rownames(raw_counts)[!gsg$goodGenes]
  
  # Print Sample Info
  if(length(removed_samples) > 0) {
    print(paste("Removing", length(removed_samples), "samples flagged by goodSamplesGenes:"))
    print(removed_samples)
  } else {
    print("No samples were flagged for removal by goodSamplesGenes.")
  }
  
  # Print Gene Info
  if(length(removed_genes) > 0) {
    print(paste("Removing", length(removed_genes), "genes flagged by goodSamplesGenes."))
  }

  # Perform removal
  gsg_counts <- raw_counts[gsg$goodGenes, gsg$goodSamples]
  
  # Update metadata
  metadata <- metadata %>% filter(SRR_ID %in% colnames(gsg_counts))
  # Re-order metadata to match the new counts columns
  metadata <- metadata[match(colnames(gsg_counts), metadata$SRR_ID), ]

  print("Data subsetting complete.")
  
} else {
  gsg_counts <- raw_counts # No changes if allOK
  # Ensure metadata is still in sync
  metadata <- metadata[match(colnames(gsg_counts), metadata$SRR_ID), ]
  print("No outliers detected by goodSamplesGenes.")
}
```

### Removing Outliers: Hierarchial Clustering

```{r clustering}
# Detect outlier samples using hierarchical clustering
htree <- hclust(dist(t(gsg_counts)), method = "average")

# Define filename based on current analysis
plot_file <- paste0("Sample_Clustering_Tree_", analysis_label, ".png")

# Save Plot
png(plot_file, width = 12, height = 9, units = "in", res = 300)
par(cex = 0.6)
par(mar = c(0, 5, 2, 0))
plot(htree, main = paste("Sample Clustering:", analysis_label),
     sub = "", xlab = "", labels = colnames(gsg_counts), hang = -1)
dev.off()

# Display Plot in Document
par(cex = 0.6)
par(mar = c(0, 5, 2, 0))
plot(htree, main = paste("Sample Clustering:", analysis_label),
     sub = "", xlab = "", labels = colnames(gsg_counts), hang = -1)
```

### Removing Outliers: PCA

```{r pca}
# Detect outlier samples using PCA
pca <- prcomp(t(gsg_counts), scale. = TRUE) # Added scaling for better normalization
pca_dat <- as.data.frame(pca$x)
pca_var <- pca$sdev^2
pca_var_percent <- round(pca_var / sum(pca_var) * 100, digits = 2)

# Plot PCA
p_pca <- ggplot(pca_dat, aes(PC1, PC2)) +
  geom_point(size = 2) +
  geom_text_repel(aes(label = rownames(pca_dat)), size = 3) +
  labs(x = paste0('PC1: ', pca_var_percent[1], ' %'),
       y = paste0('PC2: ', pca_var_percent[2], ' %'),
       title = paste("PCA Outlier Detection:", analysis_label)) +
  theme_minimal()

# Save Plot
ggsave(paste0("PCA_Outlier_Detection_", analysis_label, ".png"), 
       p_pca, width = 8, height = 6)

# Display Plot
print(p_pca)
```

### Removing Outlier Samples

```{r remove_samples}
# Remove outliers from counts
# We check if sample is in the exclusion list
cleaned_counts <- gsg_counts[, !(colnames(gsg_counts) %in% samples_to_be_excluded)]

# Remove outliers from metadata
metadata_cleaned <- metadata %>%
  filter(!SRR_ID %in% samples_to_be_excluded)

# Verification
print(paste("Original samples:", ncol(gsg_counts)))
print(paste("Final samples:", ncol(cleaned_counts)))
```

## Filtering Low Counts

We filter out genes with very low expression to reduce noise and computational load. We use a relaxed threshold (25% of samples) to ensure we retain lncRNAs that might be expressed in only one biological condition (e.g. Activated).

```{r filter}
# Create DESeq2 object
dds <- DESeqDataSetFromMatrix(
  countData = round(cleaned_counts),
  colData = metadata_cleaned,
  design = ~ 1
)

# Filter Logic: Match Differential Expression criteria
# Keep genes with counts >= 5 in at least 25% of samples
min_count <- 5
min_samples_threshold <- ceiling(ncol(cleaned_counts) * 0.25)

print(paste("Filtering genes with counts >=", min_count, "in at least", min_samples_threshold, "samples."))

keep <- rowSums(counts(dds) >= min_count) >= min_samples_threshold
dds_filtered <- dds[keep, ]

print(paste("Number of genes remaining:", nrow(dds_filtered)))
```

## Normalization (Variance Stabilization)

```{r normalize}
# Perform VST while blind to experimental groups
dds_final <- vst(dds_filtered, blind = TRUE)

# Extract and transpose the normalized counts for WGCNA (Samples as rows, Genes as columns)
norm_counts <- assay(dds_final) %>% 
  t()

print(paste("Dimensions of normalized matrix:", nrow(norm_counts), 
            "samples x", ncol(norm_counts), "genes"))
```

## Soft-Thresholding Power Selection

In this step, we test different power values to find the best setting for constructing the network. The goal is to choose a "soft power" that emphasizes strong correlations (signal) and suppresses weak correlations (noise). We look for the lowest power value where the network structure achieves a "scale-free" fit (indicated by the red line R2>0.8 in the plot).

```{r soft_threshold}
# Soft-threshold test
power <- c(c(1:10), seq(from = 12, to = 50, by = 2))

# Pick soft threshold
sft <- pickSoftThreshold(norm_counts,
                         powerVector = power,
                         networkType = "signed", # To distinguish positive and negative correlations
                         verbose = 5) # verbose = 5 provides detailed output during computation

# Store the results of the soft threshold analysis
sft_data <- sft$fitIndices

# Scale-free topology fit index
topo <- ggplot(sft_data, aes(Power, SFT.R.sq, label = Power)) +
  geom_point() +
  geom_text(nudge_y = 0.1) +
  geom_hline(yintercept = 0.9, color = 'red', linetype = 'dashed') + # Target R^2
  labs(x = 'Soft Threshold Power', y = 'Scale-Free Topology Model Fit, signed R^2') +
  theme_classic() +
  ggtitle(paste("Scale-Free Topology:", analysis_label))

# Mean connectivity
connect <- ggplot(sft_data, aes(Power, mean.k., label = Power)) +
  geom_point() +
  geom_text(nudge_y = 0.1) +
  labs(x = 'Soft Threshold Power', y = 'Mean Connectivity') +
  theme_classic() +
  ggtitle(paste("Mean Connectivity:", analysis_label))

# Display and Save
p_grid <- grid.arrange(topo, connect, nrow = 2)
ggsave(paste0("Soft_Threshold_Plots_", analysis_label, ".png"), 
       plot = p_grid, width = 9, height = 7)
```

## Network Construction and Module Detection

Using the selected soft power, we construct the weighted gene co-expression network. This step calculates the topological overlap matrix (TOM) to measure gene similarity and uses hierarchical clustering to identify "modules" of co-expressed genes.

```{r network}
# Convert matrix to numeric
norm_counts[] <- sapply(norm_counts, as.numeric)

# Temporarily store the original 'cor' function to avoid conflicts with WGCNA's 'cor'
temp_cor <- cor
cor <- WGCNA::cor

# Build the network and detect modules
bwnet <- blockwiseModules(norm_counts,
                          maxBlockSize = ncol(norm_counts), # Set to actual number of genes
                          TOMType = "signed", # Match networkType from pickSoftThreshold
                          power = soft_power, # Use the chosen soft threshold power
                          mergeCutHeight = 0.25, # Merge modules with >75% similarity
                          numericLabels = FALSE, # Assign color names to modules
                          randomSeed = 1234, # For reproducibility
                          verbose = 3) # Provides progress updates

# Restore the original 'cor' function
cor <- temp_cor
```

## Module Visualization and Eigengenes

We visualize the gene dendrogram and the corresponding module colors to verify that the clustering algorithm successfully identified distinct modules. We also extract the Module Eigengenes (MEs), which represent the "summary expression profile" of each module.

```{r eigengenes}
# Extract module eigengenes
module_eigengenes <- bwnet$MEs

# Print out a preview of the module eigengenes (each column is an eigengene for a module)
print("Preview of Module Eigengenes:")
head(module_eigengenes)

# Print Module Counts
print("Number of genes per module:")
print(table(bwnet$colors))

# Define Plotting Function (Wrapper to avoid code duplication)
plot_dendro <- function() {
  plotDendroAndColors(bwnet$dendrograms[[1]], 
                      cbind(bwnet$unmergedColors, bwnet$colors),
                      c("Unmerged", "Merged"), 
                      dendroLabels = FALSE, 
                      addGuide = TRUE, 
                      hang = 0.03,
                      guideHang = 0.05,
                      main = paste("Gene Dendrogram:", analysis_label))
}

# Save to File
png(paste0("Gene_Dendrogram_", analysis_label, ".png"), width = 12, height = 9, units = "in", res = 300)
plot_dendro()
dev.off()

# Display in Document
plot_dendro()

# The 'grey' module (if present) contains genes that do not strongly co-express with any other module.
```

## Relate Modules to Traits

We correlate the Module Eigengenes (MEs) with the `Activation_Status` trait to identify which gene modules are significantly associated with T-cell activation.

```{r traits}
# Ensure 'Activation_Status' is a factor in your cleaned metadata.
metadata_cleaned$Activation_Status <- factor(metadata_cleaned$Activation_Status,
                                                     levels = c("resting", "activated"))

# Binarize the 'Activation_Status' categorical variable into numerical columns.
activation_bin <- binarizeCategoricalColumns(metadata_cleaned$Activation_Status,
                           includePairwise = FALSE,
                           includeLevelVsAll = TRUE,
                           minCount = 1) # Include all levels

# Bind traits to sample IDs
traits <- data.frame(row.names = rownames(module_eigengenes))
traits <- cbind(traits, activation_bin)

head(traits)
```

```{r correlation}
# Define numbers of samples for correlation calculations
nSamples <- nrow(norm_counts)
nGenes <- ncol(norm_counts)

# Correlate module eigengenes and traits
# 'use = 'p'' specifies Pearson correlation
module_trait_corr <- cor(module_eigengenes, traits, use = 'p')
# Calculate the corresponding p-values for the correlations
module_trait_pvals <- corPvalueStudent(module_trait_corr, nSamples)

print("Module-Trait Correlation Matrix:")
print(round(module_trait_corr, 2))
print("Module-Trait Correlation P-values:")
print(round(module_trait_pvals, 3))
```

## Visualize Module-Trait Association

```{r heatmap}
# Merge eigengenes with traits
heatmap_data <- merge(module_eigengenes, traits, by = 'row.names')

 # Set sample IDs as row names
rownames(heatmap_data) <- heatmap_data$Row.names
heatmap_data$Row.names <- NULL

# Plot the heatmap using CorLevelPlot
corr_heat <- CorLevelPlot(heatmap_data,
                          x = colnames(activation_bin),
                          y = colnames(module_eigengenes),
                          col = c("blue1", "skyblue", "white", "pink", "red"),
                          main = paste("Module-Trait Relationships:", analysis_label))

# SAVE THE PLOT (Correct method for trellis objects)
png(paste0("Module_Trait_Heatmap_", analysis_label, ".png"), width = 8, height = 10, units = "in", res = 300)
print(corr_heat) # You must explicitly print trellis objects
dev.off()

# Display in Document
print(corr_heat)
```

## Map Genes to Modules

We create a dataframe that links every gene to its assigned module color. This mapping is the key to extracting specific gene lists for functional enrichment.

```{r map}
module_gene_mapping <- as.data.frame(bwnet$colors)
colnames(module_gene_mapping) <- "ModuleColor"
module_gene_mapping$Gene <- rownames(module_gene_mapping)

head(module_gene_mapping)
```

## Intramodular analysis: Identifying driver genes

We calculate module membership (kME) by correlating each gene's expression with the module eigengenes. High kME values indicate "hub genes" that are central drivers of the modules's biological function.

```{r driver_genes}
# Calculate kME (Gene vs Module Eigengene correlation)
module_membership <- cor(norm_counts, module_eigengenes, use = 'p')
module_membership_pvals <- corPvalueStudent(module_membership, nSamples)

print("Preview of Module Membership (kME):")
print(head(module_membership[, 1:min(5, ncol(module_membership))]))
print("Preview of Module Membership P-values:")
print(head(module_membership_pvals[, 1:min(5, ncol(module_membership_pvals))]))
```

## Calculate Gene Significance (GS)

We calculate the correlation between gene expression and the "Activated" trait. High absolute GS indicates that the gene is strongly upregulated or downregulated during activation.

```{r gene_significance}
# Define the specific trait column name
trait_col <- "data.activated.vs.all"

if (trait_col %in% colnames(traits)) {
  
  # Calculate Correlation (GS) and P-values
  gene_signif <- cor(norm_counts, traits[[trait_col]], use = 'p')
  gene_signif_pvals <- corPvalueStudent(gene_signif, nSamples)
  
  # Create Results Dataframe
  gene_signif_df <- data.frame(Gene = rownames(gene_signif), 
                               GS_Activation = gene_signif, 
                               PValue_GS = gene_signif_pvals)
  
  # Fix column names (remove the 'data.activated.vs.all' suffix from the merge)
  colnames(gene_signif_df) <- c("Gene", "GS_Activation", "PValue_GS")
  
  # Preview top genes
  print(head(gene_signif_df %>% arrange(PValue_GS), 20))
  
} else {
  stop(paste("Trait column", trait_col, "not found in traits data."))
}
```

## Create Master Dataframe

We compile all network stats (Module Color, kME, p-values, and Gene Significance) into a single master dataframe. This file serves as the primary input for downstream hub gene selection and functional enrichment.

```{r complete_dataframe}
all_gene_info <- module_gene_mapping

# Format kME table (Module Membership)
kME_df <- as.data.frame(module_membership)
# Remove "ME" prefix if present to normalize names (e.g., "MEturquoise" -> "kME_turquoise")
colnames(kME_df) <- paste0("kME_", gsub("ME", "", colnames(kME_df)))
kME_df$Gene <- rownames(kME_df)

# Format kME p-value table
p_kME_df <- as.data.frame(module_membership_pvals)
colnames(p_kME_df) <- paste0("p.kME_", gsub("ME", "", colnames(p_kME_df)))
p_kME_df$Gene <- rownames(p_kME_df)

# Merge Module Assignment + kME + Gene Significance
gene_level_data <- all_gene_info %>%
  left_join(kME_df, by = "Gene") %>%
  left_join(p_kME_df, by = "Gene") %>%
  left_join(gene_signif_df, by = "Gene")

# Save complete dataset
save_filename <- paste0("gene_level_data_", analysis_label, ".rds")
saveRDS(gene_level_data, file = save_filename)

print("Preview of Gene Level Data:")
print(head(gene_level_data[, 1:5])) # Show first 5 cols to verify merge
```

## Extract Hub Genes

We iterate throught the significant modules (defined in the YAML params) to extract hub genes - those with high module membership (|kME| > 0.7) and significance. We also save the full list of genes in each module for background enrichment analysis.

```{r hub_genes}
# Use the modules defined in the configuration at the start of the script
print("Processing the following significant modules:")
print(significant_modules)

# Define Hub Gene Thresholds
kme_threshold <- 0.7
kme_pvalue_threshold <- 0.05

for (module_color in significant_modules) {
  print(paste("--- Processing Module:", module_color, "---"))

  # 1. Get all genes in the module
  current_module_genes <- gene_level_data %>%
    filter(ModuleColor == module_color)

  if (nrow(current_module_genes) == 0) {
    print(paste("Skipping: No genes found for module", module_color))
    next
  }

  print(paste("Total genes in", module_color, "module:",
              nrow(current_module_genes)))

  # Dynamically select the correct kME and p.kME columns for filtering
  kme_col <- paste0("kME_", module_color)
  p_kme_col <- paste0("p.kME_", module_color)

  # Check if the specific kME column exists for this module
  if (!kme_col %in% colnames(current_module_genes)) {
    print(paste("Warning: kME column", kme_col, "not found. Cannot filter by kME for this module."))
    next
  }

  # Filter for Hub Genes: high kME and low p-value
  hub_genes <- current_module_genes %>%
    filter(
      abs(.data[[kme_col]]) > kme_threshold,
      .data[[p_kme_col]] < kme_pvalue_threshold
    ) %>%
    # Select specific columns to save, including the relevant kME and p-value
    dplyr::select(Gene, ModuleColor, 
                  kME = all_of(kme_col), 
                  PValue_kME = all_of(p_kme_col),
                  GS_Activation,
                  PValue_GS) %>%
    arrange(desc(abs(kME)))

  print(paste("Number of Hub Genes in", module_color, "module:",
              nrow(hub_genes)))

  # Save Hub Genes (CSV)
  if (nrow(hub_genes) > 0) {
    write.csv(hub_genes,
              file = paste0("HubGenes_", module_color, "_", analysis_label, ".csv"),
              row.names = FALSE)
    print(paste("Saved Hub Genes to:", 
                paste0("HubGenes_", module_color, "_", analysis_label, ".csv")))
  } else {
    print(paste("No Hub Genes found for", module_color, "with current thresholds."))
  }

  # Save a simple list of ALL genes in the module
  # This list would contain all genes assigned to the module, regardless of kME or GS.
  # If you want to do enrichment on the entire module first, then this list is what you'd use.
  write.table(current_module_genes$Gene,
              file = paste0("AllGenesInModule_", module_color, "_", analysis_label, ".tsv"),
              row.names = FALSE, col.names = FALSE, quote = FALSE)
  print(paste("Saved all genes in module to:", 
              paste0("AllGenesInModule_", module_color, "_", analysis_label, ".tsv")))
}
```

## Version Control

```{r versions}
# Get loaded packages and versions
packages <- sessionInfo()$otherPkgs
pkg_info <- tibble::tibble(
  Package = names(packages),
  Version = sapply(packages, function(pkg) pkg$Version)
)

# Save to text file
readr::write_tsv(pkg_info, "WGCNA_package_versions.txt")
```
